---
title: "Finding the Closest INMET Station for Each Municipality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding the Closest INMET Station for Each Municipality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Many climate analyses require assigning each municipality to its closest meteorological station. The `nearest_stations()` function automates this process by identifying the nearest INMET station for every municipality.

This vignette demonstrates how to:

- select municipalities and INMET stations for a state;
- identify the nearest station for each municipality;
- visualize the spatial distribution of assigned stations; and
- inspect the municipality–station connections.

```{r setup packages, warning = FALSE, message = FALSE}

library(climateBR)
library(dplyr)
library(ggplot2)
library(sf)

```

# Preparing the data

The package includes datasets with Brazilian municipalities and INMET weather stations.

```{r}
data("municipality")
data("inmet_stations")

municipality_pe <- municipality |>
  filter(
    state_muni == "PE",
    code_ibge7 != "2605459" # Fernando de Noronha
  )

inmet_stations_pe <- inmet_stations |>
  filter(state_station == "PE")
```

# Finding the nearest station

Use `nearest_stations()` to identify the closest station for each municipality.

```{r}
mun_stations_pe <- nearest_stations(
  municipality = municipality_pe,
  inmet_stations = inmet_stations_pe,
  n = 1
)

head(mun_stations_pe)
```

# Municipality–station connections

```{r, echo=FALSE, warning = FALSE, message = FALSE}
tmp <- tempfile(fileext = ".rda")
on.exit(unlink(tmp), add = TRUE)

download.file(
  "https://github.com/kaiorb52/dados_municipais/raw/main/mun_24.rda",
  destfile = tmp,
  mode = "wb"
)

load(tmp)

mun_24_pe <- mun_24 |>
  filter(abbrev_state == "PE") |>
  select(code_muni, geom)

mun_stations_pe2 <- mun_24_pe |>
  left_join(
    mun_stations_pe,
    by = c("code_muni" = "code_ibge7")
  )
```

```{r fig.width=7, fig.height=6, echo=FALSE, warning = FALSE, message = FALSE}

ggplot() +
  geom_sf(
    data = mun_stations_pe2,
    aes(fill = code_wmo),
  ) +
  coord_sf(
    xlim = c(-41.25, -35),
    ylim = c(-10, -6.5)
  ) +
  theme_void() +
  guides(fill = "none")

```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
distance_lines <- mun_stations_pe |>
  left_join(
    inmet_stations_pe |>
      select(
        code_wmo,
        station_lat = lat,
        station_lon = lon
      ),
    by = "code_wmo"
  ) |>
  left_join(
    municipality_pe |>
      select(
        code_ibge7,
        mun_lat = lat,
        mun_lon = lon
      ),
    by = "code_ibge7"
  ) |>
  rowwise() |>
  mutate(
    geometry = st_sfc(
      st_linestring(
        matrix(
          c(
            mun_lon, mun_lat,
            station_lon, station_lat
          ),
          ncol = 2,
          byrow = TRUE
        )
      ),
      crs = 4326
    )
  ) |>
  st_as_sf()
```

```{r fig.width=7, fig.height=6, echo=FALSE, warning = FALSE, message = FALSE}

ggplot() +
  geom_sf(data = mun_24_pe, fill = "grey95") +
  geom_sf(data = distance_lines, linewidth = 0.2) +
  geom_point(
    data = municipality_pe,
    aes(lon, lat),
    shape = 3,
    size = 0.8
  ) +
  geom_point(
    data = inmet_stations_pe,
    aes(lon, lat),
    shape = 23,
    fill = "red",
    size = 3.5
  ) +
  coord_sf(
    xlim = c(-41.25, -35),
    ylim = c(-10, -6.5)
  ) +
  theme_void()

```

